Reproducible Research - Peer Assessment 1
========================================================

## Loading and preprocessing the data

```{r,echo=TRUE}
data<-read.csv("C:\\Users\\David\\Documents\\R\\Default_WD\\Reproducible_Research\\activity.csv")
library(data.table)
data2<-data.table(data)
data3<-data2[,mean(steps,na.rm=TRUE),by=date]
```

## What is mean total number of steps taken per day?

### 1. Histogram of daily steps

```{r fig.width=8, fig.height=6,echo=TRUE}
hist(data3$V1,col="blue",xlab="Total Daily Steps",main="Distribution of Steps Taken Daily")
```

### 2. Calculating Mean and Median of the Number of Steps

```{r,echo=TRUE}
calcmean<-mean(data3$V1,na.rm=TRUE)
calcmedian<-median(data3$V1,na.rm=TRUE)
```

The mean of the total number of steps taken per day is `r calcmean` and the median of the total number of steps taken per day is `r calcmedian`.

## What is the average daily activity pattern?

### 1. Time series plot of average steps take by the time interval

```{r fig.width=10, fig.height=6,echo=TRUE}
library(data.table)
data4<-data.table(data)
dataavg<-data4[,mean(steps,na.rm=TRUE),by=interval]
plot(dataavg$interval,dataavg$V1,xlab="Interval",type="l", ylab="Average Steps",main="Time Series Plot of Average Steps Taken by Interval")
```

### 2. Maximum steps interval

```{r,echo=TRUE}
maximumavg<-dataavg$interval[dataavg$V1==max(dataavg$V1)]
```

The interval with the maximum number of steps is `r maximumavg`.

## Imputing missing values

### 1. Calculating and Reporting the Number of Missing Values in the Dataset

```{r,echo=TRUE}
missval<-sum(is.na(data[,1]))
```

The total number of missing values in the dataset is `r missval`.

### 2. Process for imputing the missing values 

The strategy used for imputing missing values is to use the mean of valid values for that interval.

```{r,echo=TRUE}
data5<-merge(data,dataavg,by="interval")
for(i in 1:nrow(data))
  {if(is.na(data5$steps[i])){data5$steps[i]=data5$V1[i]}}
```

### 3. New dataset with missing data filled in

```{r,echo=TRUE}
data6<-data5[,!(names(data5) %in% "V1")]
library(data.table)
data7<-data.table(data6)
data8<-data7[,mean(steps,na.rm=TRUE),by=date]
```

### 4.a Histogram of daily steps

```{r fig.width=8, fig.height=6,echo=TRUE}
hist(data3$V1,col="blue",xlab="Total Daily Steps",main="Distribution of Steps Taken Daily with Imputed Data")
```

### 4.b Calculating Mean and Median of the Number of Steps

```{r,echo=TRUE}
calcmean2<-mean(data8$V1,na.rm=TRUE)
calcmedian2<-median(data8$V1,na.rm=TRUE)
```

The mean of the total number of steps taken per day is `r calcmean2` and the median of the total number of steps taken per day is `r calcmedian2`.

Imputing the missing data has a negligible effect on the overall results. The same conclusions will be drawn with or without imputing the missing data.

## Are there differences in activity Patterns between Weekdays and Weekends?**

**5.1. Creating a New Factor Variable**
```{r,echo=TRUE}
data6<-data5[,!(names(data5) %in% "V1")]
data6$date <- strptime(as.character(data6$date), "%Y-%m-%d")
data6["flag"]<-NA
data6["flag"]<-weekdays(data6$date)
data6$flag[data6$flag == "Monday"]="Weekday"
data6$flag[data6$flag == "Tuesday"]="Weekday"
data6$flag[data6$flag == "Wednesday"]="Weekday"
data6$flag[data6$flag == "Thursday"]="Weekday"
data6$flag[data6$flag == "Friday"]="Weekday"
data6$flag[data6$flag == "Saturday"]="Weekend"
data6$flag[data6$flag == "Sunday"]="Weekend"
data6$flag<-as.factor(data6$flag)
```

### 2. Panel plot of time series data
```{r fig.width=8, fig.height=10,echo=TRUE}
par(mfrow=c(2,1))
weekday<-subset(data6,data6$flag=="Weekday")
weekend<-subset(data6,data6$flag=="Weekend")
library(data.table)
weekday_table<-data.table(weekday)
avg_weekday<-weekday_table[,mean(steps,na.rm=TRUE),by=interval]
avg_weekday<-avg_weekday[order(avg_weekday$interval)]
plot(as.numeric(avg_weekday$interval),avg_weekday$V1,xlab="Interval",type="l", 
     ylab="Avg Daily Steps",main="Weekdays")
library(data.table)
weekend_table<-data.table(weekend)
avg_weekend<-weekend_table[,mean(steps,na.rm=TRUE),by=interval]
avg_weekend<-avg_weekend[order(avg_weekend$interval)]
plot(as.numeric(avg_weekend$interval),avg_weekend$V1,xlab="Interval",type="l", 
    ylab="Avg Daily Steps",main="Weekends")
```

Activity patterns differ from weekdays to weekends, with more evening and late night activity during the weekends. However, the peak weekday activity is higher than the peaks in the weekends.